Correcting for bias in high-throughput sequencing data

### 2011-09-25 seqbias-paper.tex
description of qpcr analysis

We applied our procedure (labeled ``BN'') as well as those of
\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
are implemented in the R packages \texttt{mseq} and \texttt{Genominator},
respectively. The procedures were applied to four publicly available data
sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
an unpublished data set of our own (Table \ref{tab:datasets}).

@@

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
of long, highly expressed exons from the remaining chromosomes. In particular, for
each reference sequence, beginning with the set of exons annotated by Ensembl
release 60 \cite{Hubbard2009}, we removed any exons with known alternate splice
sites, then chose the top 1000 exons by read count, restricting ourselves those
at least 100 nt long.

@@

All data sets were mapped using Bowtie \citep{Langmead2009} against,
respectively, the hg19, mm9, and panTro2 genome assemblies obtained from the
UCSC Genome Browser \citep{Karolchik2008}.

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform.
%% TODO: refer to supplement for figure of bias

To evaluate the efficacy of each of the bias correction methods considered, we
counted reads overlapping each gene, defining the gene by the union of every
transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
100 additional bases from both ends of the gene. Counts were then normalized
(divided) by the length of these genes. We then removed any genes with a read
count less than 10, or that did not correspond to a unique TaqMan probe, leaving
592 genes.

%% TODO: something goes here
After adjusting read counts according to the predicted sequence bias, we
computed the Pearson correlation between log read counts and log TaqMan
expression values, which are averaged across three replicates.

%\begin{figure}
%\centerline{\includegraphics[width=0.4\textwidth]{pcrcor.eps}}
%\caption{
    %Pearson correlation is computed betweer TaqMan values and (bias-corrected)
    %read counts, on a logarithmic scale.
    %%%TODO
    %Error bars show a 95\% confidence interval, estimated using the Fisher
    %transformation, as computed by the R function \texttt{cor.test}.
%}
%\label{fig:pcrcor}
%\end{figure}

\begin{table}

\processtable{
    Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
    values.\label{tab:pcr}}{
    \centerline{
\begin{tabular*}{0.4\textwidth}{lrl}
\toprule
Method & Correlation & \\ \midrule
Unadjusted &  0.6665 & $\ast\ast$ \\
7mer &  0.6677 & $\ast\ast$ \\
GLM & 0.6854 & $\ast\ast$ \\
MART & 0.6921 & $\ast$ \\
BN & \textbf{0.7016}
\end{tabular*}}

}
{
We estimated the statistical significance of the improvement in correlation
using the BN method over the other methods using a simple boostrap procedure. A
bootstrap sample is formed by sampling, with replacement, 592 genes from the
original set of the same size. The correlation is then computed for this set,
using the adjusted count from each method. We repeated this procedure one
million times, and counted the number of times each of the competing method achieves a
higher correlation than the BN method.  Those marked with a sigle asterisk
achieved a higher correlation fewer than 1000 times, resulting in a p-value of
less than $10^{-3}$ Those marked with two asterisks achieved a higher
correlation in none of the bootstrap samples, indicating a p-value less than
$10^{-6}$.
}
\end{table}

>>>>

To evaluate the first two assumption, we applied our procedure (labeled ``BN'')
as well as those of \citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010}
(``7mer''), which are implemented in the R packages \texttt{mseq} and
\texttt{Genominator}, respectively, to four publicly available data sets
\citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as an
unpublished data set of our own (Table \ref{tab:datasets}).

@@

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
of long, highly expressed exons from the remaining chromosomes. In particular,
for each reference sequence, beginning with the set of exons annotated by
Ensembl release 60 \cite{Hubbard2009}, we removed any exons with known alternate
splice sites, then chose the top 1000 exons by read count, restricting ourselves
to those at least 100 nt long.

@@

All data sets were mapped using Bowtie \citep{Langmead2009} against,
respectively, the hg19, mm9, rheMac2, and panTro2 genome assemblies obtained
from the UCSC Genome Browser \citep{Karolchik2008}.

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform (see, Supplementary Section 5).

To evaluate the efficacy of each of the bias correction methods considered, we
counted reads overlapping each gene, defining the gene by the union of every
transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
100 additional bases from both ends of the gene. Counts were then normalized by
dividing by the length of these genes. We then removed any genes with a read
count less than 10, or that did not correspond to

Each method was trained in a manner identical to that used in the analysis of
section 3.1 and 3.2, but without restricting the training data to the first
eight chromosomes.  After adjusting read counts according to the predicted
sequence bias, we computed the Pearson correlation between log read counts and
log TaqMan expression values, which are averaged across three replicates. These
correlations are listed in Table \ref{tab:pcr}.

\begin{table}
\processtable{
    Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
    values.\label{tab:pcr}}{
    \centerline{
\begin{tabular*}{0.4\textwidth}{lrl}
\toprule
Method & Correlation & \\ \midrule
Unadjusted &  0.6665 & $\ast\ast$ \\
7mer &  0.6677 & $\ast\ast$ \\
GLM & 0.6854 & $\ast\ast$ \\
MART & 0.6921 & $\ast$ \\
BN & \textbf{0.7016}
\end{tabular*}}
}
{
We estimated the statistical significance of the improvement in correlation
using the BN method over the other methods using a simple boostrap procedure. A
bootstrap sample is formed by sampling, with replacement, 592 genes from the
original set of the same size. The correlation is then computed for this set,
using the adjusted count from each method. We repeated this procedure one
million times, and counted the number of times each of the competing method achieves a
higher correlation than the BN method.  Those marked with a sigle asterisk
achieved a higher correlation fewer than 1000 times, resulting in a p-value
less than $10^{-3}$.  Those marked with two asterisks achieved a higher
correlation in none of the bootstrap samples, indicating a p-value less than
$10^{-6}$.
}
\end{table}
